Full Dissertation

### 2024-04-05 Sections/Discussion.qmd
update with conclusion

Optional: - emphasize that both tasks involve extrapolation - it's just emphasized more in Project 2, and project 2 is designed to more closely mirror tasks used in the function learning literature.

## Project comparison

@@

Task Structure - hitting target vs. hitting wall with correct force

Feedback - relation between feedback and correction - usefulness of feedback was part of what makes HTW different from HTT. And in general in visuomotor learning tasks, feedback usually is not quite as precise as it is in HTW. - consistency of HTW finding across continuous vs. ordinal makes it seem like this wasn't the sole determinative factor.

\ \ \ \ \

## General Discussion

### Summary

Across both projects, we investigated the influence of training variability on learning and generalization in computerized visuomotor skill learning, and function learning tasks. In project 1 (HTT), experiments 1 and 2 demonstrated that varied training led to superior testing performance compared to constant training. In Experiment 1, the varied group even outperformed the constant group even when testing from the constant groups trained position. In contrast, Project 2 (HTW) found the opposite pattern - the varied training groups exhibited poorer performance than the constant groups, both in terms of final training accuracy and the ability to accurately extrapolate to novel velocity bands during testing. This detrimental effect of variability was observed across three experiments, with the exception of the reverse order condition in Experiment 3, where the varied group was able to match the constant group's performance.

Both projects also contributed computational modeling componenents.

@@

Although the HTT and HTW tasks are similiar in requiring participants to learn how to correctly launch projectiles,  they nevertheless differ across numerous dimensions that may be relevant to the opposing patterns observed in the two projects (see @tbl-task-diff provides for a detailed comparison of the two tasks).

@@

The HTT and HTW tasks also differed in terms of general task complexity. The HTT task was designed to mimic projectile launching tasks commonly employed in visuomotor learning studies, and the parabolic trajectories necessary to strike the target in HTT were sensitive to both the x and y dimensions of the projectiles velocity (and to a lesser extent, the position within the launching box at which the ball was released). Conversely the HTW task was influenced to a greater extent by the tasks commonly utilized in the function learning literature, wherein the correct output respones are determined by a single input dimension. *In HTW, the task space is also almost perfectly smooth, at least for the continuous feedback subjects, if they throw 100 units too hard, they'll be told that they were 100 units too hard. Whereas in HTT,  it was possible to produce xy velocity combinations that were technically closer to the empirical solution space than other throws, but which resulted in worse feedback due to striking the barrier.*

@@

@@

Rework the section on "Differences between the two Projects" include more speculation on how the perceptual distinctiveness of the task conditions may have memory or attentional consequences that relate to when training variation is useful. In particular discussing connections to ideas from instance based, or exemplar-based memory models of learning (e.g. works of Gordan Logan; Nosofsky; Crump, Hintzman, Estes etc.)

### Empirical Results and Modeling Summary

This summary encapsulates the empirical investigations and computational modeling efforts within the Hit The Target (HTT) and Hit The Wall (HTW) projects, emphasizing the nuanced role of training variability on visuomotor learning and generalization.

**Project 1 (HTT)** explored spatial training variability's effects. **Empirical results** from Experiments 1 and 2 consistently showed that varied training enhances transfer performance, despite initial training challenges. The **Instance-Based Generalization and Adaptation (IGAS) model** was developed to interpret these results, offering a computational perspective that underscores the importance of specific training experiences in shaping generalization. By quantifying similarity across varied training instances, the IGAS model successfully explained why varied training could lead to superior transfer outcomes, validating the hypothesis that broader training experiences foster a more adaptable skill set.

**Project 2 (HTW)** focused on velocity training variability and its influence on function learning tasks. **Empirical findings** across three experiments demonstrated varied training's differential effects on extrapolation capabilities, with notable variations depending on the specific conditions and feedback types. The **Associative Learning Model (ALM) and EXemplar-based Associative Memory (EXAM) models** were applied to these findings, elucidating the mechanisms behind the observed patterns. ALM highlighted the role of associative mechanisms in learning from varied velocity training, while EXAM offered insights into how exemplar-based memory processes might support extrapolation. Together, these models provided a comprehensive understanding of how varied training influences learning and extrapolation, showcasing the complexity of underlying cognitive processes.

**Comparative analysis** between HTT and HTW, supported by the IGAS, ALM, and EXAM models, revealed the multifaceted nature of training variability's impact. While empirical data underscored the beneficial effects of variability on learning and generalization, the models offered a deeper understanding of the cognitive and memory-related mechanisms at play. Specifically, they suggested that the effectiveness of varied training might depend on the task's perceptual features and the cognitive operations required for task performance.

**Conclusions** drawn from both the empirical and modeling efforts highlight the critical role of training variability in enhancing learning outcomes. The integration of computational models with empirical data not only validates the observed benefits of varied training but also advances our understanding of the cognitive foundations underlying these effects. Future research directions may include refining these models to capture additional nuances of learning processes and exploring their applicability to other learning domains. This holistic approach, combining empirical research with computational modeling, promises to yield rich insights into the complexities of learning and generalization.

>>>>

@@

### Empirical and Modeling Summary

Across both projects, we investigated the influence of training variability on learning and generalization in computerized visuomotor skill learning, and function learning tasks. In project 1 (HTT), experiments 1 and 2 demonstrated that varied training led to superior testing performance compared to constant training. In Experiment 1, the varied group even outperformed the constant group even when testing from the constant groups trained position. In contrast, Project 2 (HTW) found the opposite pattern - the varied training groups exhibited poorer performance than the constant groups, both in terms of training accuracy, accuracy in extrapolation testing,  and, in a subset of the experiments, the varied group showed a diminished ability to discriminate between bands. This detrimental effect of variability was observed across three experiments, with the exception of the reverse order condition in Experiment 3, where the varied group was able to match the constant group's performance.

Both projects also included computational modeling componenents. In Project 1, the IGAS model was introduced as a means of addressing the lack of control for similarity between training and testing conditions common to previous work in the "benefits of variability" literature. The IGAS model provides a theoretically motivated method of quantifying the similarity between training experience and testing conditions. The resulting similarity metric (i.e. our 1c-similarity) is shown to be a significant predictor of testing performance on its own, and when added as a covariate to the statistical model used to compare the constant and varied training groups. We then showed the group-level effect of training variability on testing performance can be accounted for with the additional assumption that training variability influences the generalization gradient. The contribution of the IGAS model was thus twofold:  1) providing a theoretically justifiable method of quantifying/controlling for similarity between training and testing, and 2) demonstrating the viability of a flexible-similarity based generalization account for the empirically observed benefit of variability in our task. Although similar approaches have been employed in other domains, both contributions are novel additions to the large body of research assessing the effect of constant vs. varied training manipulations in visuomotor skill tasks.

Although theoretically motivated, the IGAS model of Project 1 is best categorized as a descriptive measurement-model. Sufficient to account for group differences, but lacking the machinery necessary to provide a full process-level account of how the empirical quantities of interest are generated. In contrast, Project 2 (HTW) implemented a more robust computational modeling approach, implementing and comparing full process models (ALM & EXAM), capable of generating predictions for both the learning and testing stages of the experiment. ALM and EXAM have been used as models of function learning, cue judgement, and forecasting behavior in numerous studies over the past 25 years [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @brownUnderestimationLinearFunction2017; @kwantesItemOrderMatters2012; @kelleyComparisonModelsLearning2008; @kaneApplicationsBiasVariance2020; @vonhelversenLearningMultiplecueJudgment2010]. The present work presents the first application of these models to to the study of training variability in a visuomotor function learning task. We fit both models to individual participant data, using a form of simulation-based bayesian parameter estimation that allowed us to generate and compare the full posterior predictive distributions of each model. EXAM provided the best overall account of the testing data, and the advantage of EXAM over ALM was significantly greater for the constant group. Notably, EXAM captured the constant groups' ability to extrapolate linearly to novel velocity bands, despite receiving training from only a single input-output pair. This finding suggests that EXAM's linear extrapolation mechanism, combined with the assumption of prior knowledge about the origin point (0, 0), was sufficient to account for the constant groups' accurate extrapolation performance. Such findings may offer a preliminary suggestion that experience with a more variable set of training examples  may be detrimental to performance in simple extrapolation tasks.

@@

The HTT and HTW tasks differ across numerous dimensions that may be relevant to the opposing patterns observed in the two projects (see @tbl-task-diff provides for a detailed comparison of the two tasks).

@@

The HTT and HTW tasks also differed in terms of general task complexity. The HTT task was designed to mimic projectile launching tasks commonly employed in visuomotor learning studies, and the parabolic trajectories necessary to strike the target in HTT were sensitive to both the x and y dimensions of the projectiles velocity (and to a lesser extent, the position within the launching box at which the ball was released). Conversely the HTW task was influenced to a greater extent by the tasks commonly utilized in the function learning literature, wherein the correct output respones are determined by a single input dimension. In HTW,the relationship between feedback and optimal behavioral adjustment is also almost perfectly smooth, if participants produce a throw that is 100 units too hard, they'll be told that they were 100 units away from the target band. Whereas in HTT, the presence of the barrier in introduces irregularities in the task space. Even throws close to the solution space might result in failure, creating a less predictable learning environment.

@@

### Conclusion

In summary, this dissertation provides a comprehensive examination of the effects of training variability on learning and generalization in visuomotor and function learning tasks. The contrasting results obtained from the Hit The Target (HTT) and Hit The Wall (HTW) tasks underscore the complexity inherent to the longstanding pedagogical and scientific goal of identifying training manipulations that consistently benefit learning and generalization. Moreover, through the development and application of computational models, we provide novel theoretical accounts for both the beneficial and detrimental effects of training variability observed in our experiments. These findings highlight the importance of considering task characteristics when designing experiments intended to assess the influence of training interventions, and demonstrate the value of combining empirical and computational modeling approaches to uncover the cognitive mechanisms that support learning and generalization. Future research should continue to investigate the complex interplay between task demands, training manipulations, and individual differences, with the ultimate goal of optimizing educational and training outcomes across a wide range of domains.

@@

### 2024-04-05 Sections/full.qmd
update with conclusion

## General Discussion

@@

### Empirical Results Summary

Across two projects, we investigated the influence of training variability on learning and generalization in visuomotor tasks. In Project 1 (HTT), we consistently found that varied training led to superior performance relative to constant training, even when testing at positions that were identical to the constant group's training. In contrast, Project 2 (HTW) found the opposite pattern - the varied training groups exhibited poorer performance than the constant groups, both in terms of final training accuracy and the ability to accurately extrapolate to novel velocity bands during testing. This detrimental effect of variability was observed across three experiments, with the exception of the reverse order condition in Experiment 3, where the varied group was able to match the constant group's performance.

@@

The HTT and HTW tasks also differed in terms of general task complexity. The HTT task was designed to mimic projectile launching tasks commonly employed in visuomotor learning studies, and the parabolic trajectories necessary to strike the target in HTT were sensitive to both the x and y dimensions of the projectiles velocity (and to a lesser extent, the position within the launching box at which the ball was released). Conversely the HTW task was influenced to a greater extent by the tasks commonly utilized in the function learning literature, wherein the correct output respones are determined by a single input dimension.

@@

>>>>

@@

### Empirical and Modeling Summary

Across both projects, we investigated the influence of training variability on learning and generalization in computerized visuomotor skill learning, and function learning tasks. In project 1 (HTT), experiments 1 and 2 demonstrated that varied training led to superior testing performance compared to constant training. In Experiment 1, the varied group even outperformed the constant group even when testing from the constant groups trained position. In contrast, Project 2 (HTW) found the opposite pattern - the varied training groups exhibited poorer performance than the constant groups, both in terms of training accuracy, accuracy in extrapolation testing,  and, in a subset of the experiments, the varied group showed a diminished ability to discriminate between bands. This detrimental effect of variability was observed across three experiments, with the exception of the reverse order condition in Experiment 3, where the varied group was able to match the constant group's performance.

Both projects also included computational modeling componenents. In Project 1, the IGAS model was introduced as a means of addressing the lack of control for similarity between training and testing conditions common to previous work in the "benefits of variability" literature. The IGAS model provides a theoretically motivated method of quantifying the similarity between training experience and testing conditions. The resulting similarity metric (i.e. our 1c-similarity) is shown to be a significant predictor of testing performance on its own, and when added as a covariate to the statistical model used to compare the constant and varied training groups. We then showed the group-level effect of training variability on testing performance can be accounted for with the additional assumption that training variability influences the generalization gradient. The contribution of the IGAS model was thus twofold:  1) providing a theoretically justifiable method of quantifying/controlling for similarity between training and testing, and 2) demonstrating the viability of a flexible-similarity based generalization account for the empirically observed benefit of variability in our task. Although similar approaches have been employed in other domains, both contributions are novel additions to the large body of research assessing the effect of constant vs. varied training manipulations in visuomotor skill tasks.

Although theoretically motivated, the IGAS model of Project 1 is best categorized as a descriptive measurement-model. Sufficient to account for group differences, but lacking the machinery necessary to provide a full process-level account of how the empirical quantities of interest are generated. In contrast, Project 2 (HTW) implemented a more robust computational modeling approach, implementing and comparing full process models (ALM & EXAM), capable of generating predictions for both the learning and testing stages of the experiment. ALM and EXAM have been used as models of function learning, cue judgement, and forecasting behavior in numerous studies over the past 25 years [@deloshExtrapolationSineQua1997; @mcdanielPredictingTransferPerformance2009; @brownUnderestimationLinearFunction2017; @kwantesItemOrderMatters2012; @kelleyComparisonModelsLearning2008; @kaneApplicationsBiasVariance2020; @vonhelversenLearningMultiplecueJudgment2010]. The present work presents the first application of these models to to the study of training variability in a visuomotor function learning task. We fit both models to individual participant data, using a form of simulation-based bayesian parameter estimation that allowed us to generate and compare the full posterior predictive distributions of each model. EXAM provided the best overall account of the testing data, and the advantage of EXAM over ALM was significantly greater for the constant group. Notably, EXAM captured the constant groups' ability to extrapolate linearly to novel velocity bands, despite receiving training from only a single input-output pair. This finding suggests that EXAM's linear extrapolation mechanism, combined with the assumption of prior knowledge about the origin point (0, 0), was sufficient to account for the constant groups' accurate extrapolation performance. Such findings may offer a preliminary suggestion that experience with a more variable set of training examples  may be detrimental to performance in simple extrapolation tasks.

@@

The HTT and HTW tasks also differed in terms of general task complexity. The HTT task was designed to mimic projectile launching tasks commonly employed in visuomotor learning studies, and the parabolic trajectories necessary to strike the target in HTT were sensitive to both the x and y dimensions of the projectiles velocity (and to a lesser extent, the position within the launching box at which the ball was released). Conversely the HTW task was influenced to a greater extent by the tasks commonly utilized in the function learning literature, wherein the correct output respones are determined by a single input dimension. In HTW,the relationship between feedback and optimal behavioral adjustment is also almost perfectly smooth, if participants produce a throw that is 100 units too hard, they'll be told that they were 100 units away from the target band. Whereas in HTT, the presence of the barrier in introduces irregularities in the task space. Even throws close to the solution space might result in failure, creating a less predictable learning environment.

@@

### Conclusion

In summary, this dissertation provides a comprehensive examination of the effects of training variability on learning and generalization in visuomotor and function learning tasks. The contrasting results obtained from the Hit The Target (HTT) and Hit The Wall (HTW) tasks underscore the complexity inherent to the longstanding pedagogical and scientific goal of identifying training manipulations that consistently benefit learning and generalization. Moreover, through the development and application of computational models, we provide novel theoretical accounts for both the beneficial and detrimental effects of training variability observed in our experiments. These findings highlight the importance of considering task characteristics when designing experiments intended to assess the influence of training interventions, and demonstrate the value of combining empirical and computational modeling approaches to uncover the cognitive mechanisms that support learning and generalization. Future research should continue to investigate the complex interplay between task demands, training manipulations, and individual differences, with the ultimate goal of optimizing educational and training outcomes across a wide range of domains.

\ \ \ \
